simodels: A package for Spatial Interaction Modelling

### 2024-08-02 paper/paper.qmd
Diagnose reason of weird result (inner not left join), close #34

bibliography: bibliography.bib
execute: 
  messages: false
  warning: false 
---

<!-- ## Introduction: Count data regression in R {#sec-intro}

@@

More recent innovations in SIMs including the 'radiation model' @simini_universal_2012.
See @lenormand_systematic_2016 for a comparison of alternative approaches.

Before describing the functions in the package, it's worth implementing SIMs from first principles, to gain an understanding of how they work.
The code presented below was written before the functions in the [simodels]{.pkg} package were developed, building on @dennett_modelling_2018.
The aim is to demonstrate a common way of running SIMs, in a for loop, rather than using vectorised operations used in [simodels]{.pkg}.
We will lust the following packages for data manipulation and manipulation, using only base [R]{.proglang} functions for the SIM calculations.

```{r}
#| label: load-packages
library(tmap)
library(dplyr)
library(ggplot2)
```

- `si_zones`: a dataset of zones with population data
- `si_centroids`: a dataset of zone centroids
- `si_od_census`: a dataset of Origin-Destination (OD) data for evaluating the SIMs

We'll use the [od](.pkg) package to convert the geographic data to OD data.
We set the `length` column to the length of the line between the centroids of the zones, in km, and set the coordinate reference system (CRS) to WGS84 (EPSG:4326).

```{r}
od_df = od::points_to_od(centroids)
od_sfc = od::odc_to_sfc(od_df[3:6])
sf::st_crs(od_sfc) = 4326
od_df$length = sf::st_length(od_sfc)
od_df = od_df |>
  transmute(
    O, D, length = as.numeric(length) / 1000,
    flow = NA, fc = NA
  )
od_df = sf::st_sf(od_df, geometry = od_sfc, crs = 4326)
```

```{r unconstrained}
beta = 0.3
for(i in seq(nrow(zones))) {
  for(j in seq(nrow(zones))) {
    O = zones$all[i]
    n = zones$all[j]
    ij = which(od_df$O == zones$geo_code[i] & od_df$D == zones$geo_code[j])
    od_df$fc[ij] = exp(-beta * od_df$length[ij])
    od_df$flow[ij] = O * n * od_df$fc[ij]
  }
}
```

We can make this production constrained as follows:

```{r constrained}
od_dfj = left_join(
  od_df,
  zones |> select(O = geo_code, all) |> sf::st_drop_geometry()
)
od_dfj = od_dfj |> 
  group_by(O) |> 
  mutate(flow_constrained = flow / sum(flow) * first(all)) |>
  ungroup()
sum(od_dfj$flow_constrained) == sum(zones$all)
od_top = od_dfj |> 
  filter(O != D) |> 
  arrange(desc(flow_constrained)) |>
  slice(seq(2000)) |>
  arrange(flow_constrained)

>>>>

bibliography: bibliography.bib
execute: 
  messages: false
  warning: false 
  echo: false
---

More recent innovations in SIMs including the 'radiation model' @simini_universal_2012.
See @lenormand_systematic_2016 for a comparison of alternative approaches.
The purpose of the package is not to implement all available models, but to provide a flexible framework to speed-up the process of implementing SIMs in [R]{.proglang}.
The package supports a wide range of models and can be used to fit unconstrained, production-constrained, attraction-constrained, and doubly-constrained models.
Furthermore, the package works equally well for 'bipartite' SIMs, in which origin and destination points are different (Hasova et al. [2022](https://lenkahas.com/files/preprint.pdf)).

Before describing the functions in the package, it's worth implementing SIMs from first principles, to gain an understanding of how they work and to highlight the benefits of using the package.
The code presented below was written before the functions in the [simodels]{.pkg} package were developed, building on @dennett_modelling_2018.
The aim is to demonstrate a common way of running SIMs, in a for loop, rather than using vectorised operations used in [simodels]{.pkg}.
In this section we the [tmap]{.pkg} and [tidyverse]{.pkg} packages for data manipulation and visualisation, using only base [R]{.proglang} functions for the SIM calculations (see the paper's source code for details).

```{r}
#| label: load-packages
library(tmap)
library(tidyverse)
library(patchwork)
devtools::load_all()
```

- `si_zones`: a dataset of zones with population data
- `si_centroids`: a dataset of zone centroids
- `si_od_census`: a dataset of Origin-Destination (OD) data for evaluating the SIMs
- `zones_york`: a dataset of zones in York
- `destinations_york`: a dataset of school locations in York with information on number of pupils attending, enabling destination-constrained and doubly-constrained SIMs to be implemented

We'll use the [od](.pkg) package to convert the geographic data to OD data.
We set the `length` column to the length of the line between the centroids of the zones, in km, and set the coordinate reference system (CRS) to WGS84 (EPSG:4326), as follows:

```{r}
#| echo: true
od_df = od::points_to_od(centroids)
od_sfc = od::odc_to_sfc(od_df[3:6])
sf::st_crs(od_sfc) = 4326
od_df$length = sf::st_length(od_sfc)
od_df = od_df |>
  transmute(
    O, D, length = as.numeric(length) / 1000,
    flow = NA, fc = NA
  )
od_df = sf::st_sf(od_df, geometry = od_sfc, crs = 4326)
```

```{r}
#| label: unconstrained
#| echo: true
beta = 0.3
for(i in seq(nrow(zones))) {
  for(j in seq(nrow(zones))) {
    O = zones$all[i]
    n = zones$all[j]
    ij = which(od_df$O == zones$geo_code[i] & od_df$D == zones$geo_code[j])
    od_df$fc[ij] = exp(-beta * od_df$length[ij])
    od_df$flow[ij] = O * n * od_df$fc[ij]
  }
}
```

We can make this production constrained by dividing the estimated interaction by the difference between the sum of the estimated interactions and the sum of the observed interactions per origin.

```{r}
#| label: constrained
#| include: false
od_dfj = left_join(
  od_df,
  zones |> select(O = geo_code, all) |> sf::st_drop_geometry()
)
od_dfj = od_dfj |> 
  group_by(O) |> 
  mutate(flow_constrained = flow / sum(flow) * first(all)) |>
  ungroup()
sum(od_dfj$flow_constrained) == sum(zones$all)
od_top = od_dfj |> 
  filter(O != D) |> 
  arrange(desc(flow_constrained)) |>
  slice(seq(2000)) |>
  arrange(flow_constrained)

```{r}
#| label: inputs-york
g1 = zones_york |>
  ggplot() +
  geom_sf() +
  labs(title = "Zones")
g2 = destinations_york |>
    ggplot() +
    geom_sf() +
    labs(title = "Destinations")
g1 + g2
```

Before we run any models let's compare the total number of pupils in the zones dataset and the destinations dataset (they should be the same):

```{r}
#| label: compare-totals
#| echo: true
zone_overestimate_factor = 
  (sum(zones_york$f0_to_15) + sum(zones_york$m0_to_15)) /
    sum(destinations_york$n_pupils)
zone_overestimate_factor
```

As one would expect, the total number of pupils in the zones dataset is a bit bigger than the total number of pupils in the destinations dataset: not all people aged 0 to 15 go to school, especially those under school age.
To tackle this issue we'll create a new variables called `pupils_estimated` in the zones dataset, which is the sum of the number of pupils in the zones dataset and the number of pupils in the destinations dataset.

```{r}
#| label: add-pupils-estimated
#| echo: true
zones_york = zones_york |>
  dplyr::mutate(
    pupils_estimated = (f0_to_15 + m0_to_15) / zone_overestimate_factor
  )
```

After the adjustment shown above, the totals in the origin and destination columns should be the same:

```{r}
#| label: compare-totals-after
#| echo: true
sum(zones_york$pupils_estimated)
sum(destinations_york$n_pupils)
```

Based on these inputs the `si_to_od()` function generates the OD data, as shown below (note: 2 versions are created, one with a maximum distance constraint for speed of processing, important when working with large datasets).

```{r}
#| echo: true
max_dist = 5000 # meters
od_from_si_full = simodels::si_to_od(zones_york, destinations_york)
od_from_si = simodels::si_to_od(zones_york, destinations_york, max_dist = max_dist)
```

```{r}
#| label: plot-od-all
#| layout-ncol: 2
m1 = od_from_si_full |>
  ggplot() +
  geom_sf(alpha = 0.1)
m2 = od_from_si |>
  ggplot() +
  geom_sf(alpha = 0.1)
m1
m2
```

The output OD dataset has column names taken from both the origin and destination datasets, with the following column names:

```{r}
names(od_from_si)
```

Let's run a simple model:

```{r}
#| label: simple-model-york
#| echo: true
gravity_model = function(beta, d, m, n) {
  m * n * exp(-beta * d / 1000)
} 
# perform SIM
od_res = simodels::si_calculate(
  od_from_si,
  fun = gravity_model,
  d = distance_euclidean,
  m = origin_pupils_estimated,
  n = destination_n_pupils,
#   constraint_production = origin_all,
  beta = 0.9
  )
```

We'll make one adjustment to the output dataset, renaming the `interaction` column to `trips`, and setting the total number of trips to be the same as the total number of pupils in the destinations dataset:

```{r}
#| label: adjust-output-york
#| echo: true
interaction_overestimate_factor = sum(destinations_york$n_pupils) / sum(od_res$interaction)
od_res = od_res |>
  dplyr::mutate(
    trips = interaction * interaction_overestimate_factor
  )
```

We can assess the model fit at three levels: the origin level (number of students departing from each zone), the destination level (the number arriving at each school in the input dataset) and the origin-destination level.

```{r}
#| label: r-squared-york
res_o = od_res |>
  group_by(O) |>
  summarise(
    Observed = first(origin_pupils_estimated),
    Modelled = sum(trips),
    Type = "Origin"
  )
res_d = od_res |>
  group_by(D) |>
  summarise(
    Observed = first(destination_n_pupils),
    Modelled = sum(trips),
    Type = "Destination"
  )
res_combined = bind_rows(res_o, res_d) |>
  # Create ordered factor with types:
  mutate(
    Type = factor(Type, levels = c("Origin", "Destination"))
  )
g_combined = res_combined |>
  ggplot() +
    geom_point(aes(x = Observed, y = Modelled)) +
    geom_smooth(aes(x = Observed, y = Modelled), method = "lm") +
    facet_wrap(~Type, scales = "free") +
    labs(
      title = "Model fit at origin, destination and OD levels (unconstrained)",
      x = "Observed",
      y = "Modelled"
    ) 
g_combined
```

Let's see if making the model production constrained can help:

```{r}
#| label: constrained-york
#| echo: true
od_res_constrained = simodels::si_calculate(
  od_from_si,
  fun = gravity_model,
  d = distance_euclidean,
  m = origin_pupils_estimated,
  n = destination_n_pupils,
  constraint_production = origin_pupils_estimated,
  beta = 0.9
  )
```

```{r}
#| label: r-squared-constrained
res_o = od_res_constrained |>
  group_by(O) |>
  summarise(
    Observed = first(origin_pupils_estimated),
    Modelled = sum(interaction),
    Type = "Origin"
  )
res_d = od_res_constrained |>
  group_by(D) |>
  summarise(
    Observed = first(destination_n_pupils),
    Modelled = sum(interaction),
    Type = "Destination"
  )
res_combined = bind_rows(res_o, res_d) |>
  mutate(
    Type = factor(Type, levels = c("Origin", "Destination"))
  )
res_combined |>
  ggplot() +
    geom_point(aes(x = Observed, y = Modelled)) +
    geom_smooth(aes(x = Observed, y = Modelled), method = "lm") +
    facet_wrap(~Type, scales = "free") +
    labs(
      title = "Model fit at origin, destination and OD levels (production constrained)",
      x = "Observed",
      y = "Modelled"
    )
```

<!--

From paper-writing docs:
 ## Introduction: Count data regression in R {#sec-intro}

@@
